#Machine Learning Course Project
Jean Rafael Angeles
June 21, 2015

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).



## Preparing required packages

```{r warning=FALSE}
library(caret)
library(randomForest)
library(rpart) 
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(kernlab)
library(doParallel)
library(nnet)
library(caTools)
library(arm)

# Set up to take advantage of multiple processing cores
Run_All_Cores <- makeCluster(detectCores())
registerDoParallel(Run_All_Cores)
```
Load the required libraries and set up to use multiple processors

## Setting up training data

```{r cache=TRUE}
# Download (if necessary) and create the csv files containing training and validation data
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
validateUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

trainCsv    <- "pml-training.csv";
validateCsv <-"pml-testing.csv"

if (!file.exists(trainCsv)) { 
    download.file(trainUrl,trainCsv)
}
training <- read.csv(trainCsv, na.strings=c("NA","#DIV/0!",""))

if (!file.exists(validateCsv)) {
    download.file(validateUrl,validateCsv)
}
validation_data <- read.csv(validateCsv, na.strings=c("NA","#DIV/0!",""))

set.seed(1234)

# Partition out training and testing data.  60% testing, 40% training
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
training_data <- training[inTrain, ] 
testing_data <- training[-inTrain, ]

# Determine which features have little or no variance, and throw them out immediately
nzv_train <- nearZeroVar(training_data, saveMetrics=TRUE)
training_data <- training_data[,nzv_train$nzv==FALSE & nzv_train$zeroVar==FALSE]
testing_data <- testing_data[,nzv_train$nzv==FALSE & nzv_train$zeroVar==FALSE]
validation_data <- validation_data[,nzv_train$nzv==FALSE & nzv_train$zeroVar==FALSE]

# The first six columns of data are irrelevant
training_data <- training_data[, -(1:6)]
testing_data <- testing_data[,-(1:6)]
validation_data <- validation_data[,-(1:6)]

# Remove any columns with more than 60% NAs.
valid_columns <- lapply(training_data, function(x) sum(is.na(x)) / length(x)) <= 0.6
training_data <- training_data[ valid_columns ]
testing_data <- testing_data[ valid_columns ]
validation_data <- validation_data[ valid_columns ]
```


## Data Processing

```{r}
#Calculate both Out of Sample Accuracy and Error for each simulation type.
outOfSampleAccuracy <- numeric()
outOfSampleError <- numeric()

# Set some training parameters
tc <- trainControl(method = "cv", number = 7,
                   verboseIter=FALSE , preProcOptions="pca", allowParallel=TRUE)

# Support Vector Machines Radial
svmr <- train(classe ~ ., data = training_data, method = "svmRadial", trControl= tc)
testingPrediction <- predict(svmr, testing_data)
accuracy <- sum(testingPrediction == testing_data$classe) / length(testingPrediction)
outOfSampleAccuracy <- c(outOfSampleAccuracy, accuracy)
outOfSampleError <- c(outOfSampleError, 1-accuracy)

# Neural Net
nnet <- train(classe ~ ., data = training_data, method = "nnet", trControl= tc, verbose=FALSE)

testingPrediction <- predict(nnet, testing_data)
accuracy <- sum(testingPrediction == testing_data$classe) / length(testingPrediction)
outOfSampleAccuracy <- c(outOfSampleAccuracy, accuracy)
outOfSampleError <- c(outOfSampleError, 1-accuracy)

# Random Forest
rf <- train(classe ~ ., data = training_data, method = "rf", trControl= tc)
testingPrediction <- predict(rf, testing_data)
accuracy <- sum(testingPrediction == testing_data$classe) / length(testingPrediction)
outOfSampleAccuracy <- c(outOfSampleAccuracy, accuracy)
outOfSampleError <- c(outOfSampleError, 1-accuracy)

# K Nearest Neighbor
knn <- train(classe ~ ., data = training_data, method = "knn", trControl= tc)
testingPrediction <- predict(knn, testing_data)
accuracy <- sum(testingPrediction == testing_data$classe) / length(testingPrediction)
outOfSampleAccuracy <- c(outOfSampleAccuracy, accuracy)
outOfSampleError <- c(outOfSampleError, 1-accuracy)

# Support Vector Machines Linear
svml <- train(classe ~ ., data = training_data, method = "svmLinear", trControl= tc)
testingPrediction <- predict(svml, testing_data)
accuracy <- sum(testingPrediction == testing_data$classe) / length(testingPrediction)
outOfSampleAccuracy <- c(outOfSampleAccuracy, accuracy)
outOfSampleError <- c(outOfSampleError, 1-accuracy)

# Bayesian GLM
bayesglm <- train(classe ~ ., data = training_data, method = "bayesglm", trControl= tc)

testingPrediction <- predict(bayesglm, testing_data)
accuracy <- sum(testingPrediction == testing_data$classe) / length(testingPrediction)
outOfSampleAccuracy <- c(outOfSampleAccuracy, accuracy)
outOfSampleError <- c(outOfSampleError, 1-accuracy)

InSampleAccuracy <- c(max(svmr$results$Accuracy),
        max(nnet$results$Accuracy),
        max(rf$results$Accuracy),
        max(knn$results$Accuracy),
        max(svml$results$Accuracy),
        max(bayesglm$results$Accuracy))

InSampleAccuracy
```
Random Forests produces the best sample accuracy for the model. 

## Out of Sample Accuracy and Out of Sample Error Results
```{r}
model <- c("SVM (radial)", "Neural Net", "Random Forest", "KNN", "SVM (Linear)", "Bayesian GLM")

performance <- data.frame(model, outOfSampleError, outOfSampleAccuracy)
performance[order(performance$outOfSampleError),]
```
Random Forest also produces the lowest Out of Sample Error. As such, Random Forest will be the chosen simulation for the 20 row validation data.

## Prediction Results

```{r}
rfPred <- predict(rf, validation_data)

pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

pml_write_files(rfPred)
```

The Random Forest predictions were able to correctly predict all the 20 validation values. The results are confirmed accurate by Coursera through the "Course Project: Submission Task" evaluation.